---
title: "Reproducible Research Assignment1"
author: "Amit Sinha"
date: "Sunday, September 14, 2014"
output: html_document
---

###Loading and preprocessing the data  
####1. Load the data and Transform
```{r, echo=TRUE}
library(stats)
activity.data <- read.csv(file = "activity.csv",header = T,sep = ",")
#rearranging column order
activity.data <- data.frame(date = activity.data$date,
                            interval = activity.data$interval,
                            steps = activity.data$steps)
```


###Mean total number of steps taken per day
```{r, echo=TRUE}
daywise.steps <- aggregate.data.frame(activity.data$steps,
                                      by = list(activity.data$date),
                                      FUN = sum,na.rm = T)
colnames(daywise.steps) <- c("date","steps")
```
####1. Make histogram  

```{r histogram,fig.height = 4}
par(mar = c(5,4,1,1),las = 1)
hist(daywise.steps$steps,breaks = 50,xlab = "Steps",ylab = "Frequency",
     col = "red",main ="A histogram of the total number of steps taken each day" )
```  

  
####2. Calculate Mean and Median  
```{r,echo=TRUE}
steps.mean <- round(mean(daywise.steps$steps),0)
steps.median <- median(daywise.steps$steps)
```

1.Mean total steps taken per day : **`r format(steps.mean,scientific = FALSE)`**  
2.Median total steps taken per day: **`r format(steps.median, scientific = FALSE)`**     
  
### Average daily activity pattern

```{r, echo=TRUE}
intervalwise.steps <- aggregate.data.frame(activity.data$steps,
                                           by = list(activity.data$interval),
                                           FUN = mean,na.rm = T)
colnames(intervalwise.steps) <- c("interval","steps")
```

####1. Make times series plot 
```{r timeseries,fig.height = 4}
par(mar = c(5,4,1,1),las = 1)
plot(x = intervalwise.steps$interval,y = intervalwise.steps$steps,
     xlab = "time(5 min interval)",ylab = "Avg steps Taken",
     type = "l",main ="Average number of steps taken at 5-min interval" )
```  

####2. Interval with maximum number of steps:
The 5 min interval starting `r intervalwise.steps[which.max(x = intervalwise.steps$steps),1]`, ending
`r intervalwise.steps[which.max(x = intervalwise.steps$steps)+1,1]` contains the maximum number of steps.


### Inputing missing values
####1. Calculating number of missing values
```{r,echo=TRUE}
missing.val <- sum(is.na(activity.data$steps))

```

  
Total number of missing values in the dataset is `r missing.val`


  
####2. Filling in all of the missing values in the dataset using mean for that 5-minute interval.

```{r,echo=TRUE}
activity.newdata = activity.data
missing.df <- activity.newdata[is.na(activity.newdata$steps),]
missing.index <- match(missing.df$interval,intervalwise.steps$interval)
activity.newdata$steps[is.na(activity.newdata$steps)] <- intervalwise.steps$steps[missing.index]

```
 
 
 
  
Old data:
```{r showtable.old, results='asis'}  
library(xtable)  
xt.old <- xtable(head(activity.data))
print(xt.old,type = "html")
```
  



  
####3. New dataset with the missing data filled in.
  
New data:
```{r showtable.new, results='asis'}  
library(xtable)  
xt.new <- xtable(head(activity.newdata))
print(xt.new,type = "html")
```


  
####3. Mean total number of steps taken per day using new data set
```{r, echo=TRUE}
daywise.steps.new <- aggregate.data.frame(activity.newdata$steps,
                                      by = list(activity.newdata$date),
                                      FUN = sum,na.rm = T)
colnames(daywise.steps.new) <- c("date","steps")
```

  
####4. Histogram using new data

```{r histogram.2,fig.height = 4}
par(mar = c(5,4,1,1),las = 1)
hist(daywise.steps.new$steps,breaks = 50,xlab = "Steps",ylab = "Frequency",
     col = "red",main ="A histogram of the total number of steps taken each day(new data)" )
```  
  
    
```{r,echo=TRUE}
steps.mean.new <- round(mean(daywise.steps.new$steps),0)
steps.median.new <- median(daywise.steps.new$steps)
```
  
    
     
1. New Mean total steps taken per day : **`r format(steps.mean.new,scientific = FALSE)`**    
2. New Median total steps taken per day: **`r format(steps.median.new,scientific = FALSE)`**       

- New mean and median are equal to each other and slightly larger in value compared to old values.
- Also, the new histogram appears to be normally distributed around mean. 


  
###Differences in activity patterns between weekdays and weekends
 
  
   
####1. Adding new factor variable with two levels - "weekday" and "weekend"  

```{r,echo=TRUE}

day.type <- weekdays(as.Date(activity.newdata$date),abbreviate = T)
day.type <- ifelse(test = day.type %in% c("Sun","Sat"),yes = "weekend","weekday")
activity.newdata$day.type <- as.factor(day.type)


daywise.steps.new.panel <- aggregate.data.frame(activity.newdata$steps,
                                      by = list(activity.newdata$interval,activity.newdata$day.type),
                                      FUN = mean,na.rm = T)
colnames(daywise.steps.new.panel) <- c("interval","day.type","steps")                                      
```
  
    
####2. Panel plot 
```{r plot.2,fig.height = 4}
xyplot(steps ~ interval | day.type,data = daywise.steps.new.panel,type = "l",layout = c(1,2),,main = "Average number of steps taken")
```


  
   
  1. During weekend, number of steps are more uniformally distributed between 800 and 2000.  
  2. During weekdays, majority of the steps are concentrated between 800 and 1000.


